A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-19 tex/codatadp.tex
more

\begin{examplex}
Let $\delta$ be a codatatype with three constructors $\const{C}, \const{D}, \const{E} :\delta \rightarrow \delta$.
Let $\Ec$ be the set
$\{y_1 \teq \const{C}( y_2 ),\;
y_2 \teq \const{D}( x ),\;
z_1 \teq \const{E}( z_2 ),\;
z_2 \teq \const{C}( z_1 ),\;
w_1 \teq \const{C}( x ),\;
w_2 \teq \const{E}( w_3 ),\;
x \not\teq y_2\}$.
After running the calculus to saturation on $\Ec$, the mapping $\Val$ contains
\[\begin{array}{r@{}c@{}r@{\;}l@{}ll@{\qquad}r@{}c@{}l@{\;}l}
\Val \ec{y_1} & {}\,=\,{} & \MU {\Varec{y_1}}. & \const{C}( & \MU {\Varec{y_2}}. & \const{D}( {\Varec{x}} ) ) &
\Val \ec{w_1} & {}\,=\,{} & \MU {\Varec{w_1}}. & \const{C}( {\Varec{x}} )      \\
\Val \ec{y_2} & = & \MU {\Varec{y_2}}. & \multicolumn{2}{@{}l@{}}{\const{D}( {\Varec{x}} )}      & &
\Val \ec{w_2} & = & \MU {\Varec{w_2}}. & \const{E}( {\Varec{w_3}} ) \\
\Val \ec{z_1} & = & \MU {\Varec{z_1}}. & \const{E}( & \MU {\Varec{z_2}}. & \const{C}( {\Varec{z_1}} ) ) &
\Val \ec{w_3} & = & {\Varec{w_3}} \\
\Val \ec{z_2} & = & \MU {\Varec{z_2}}. & \const{C}( & \MU {\Varec{z_1}}. & \const{E}( {\Varec{z_2}} ) ) &
\Val \ec{x} & = & {\Varec{x}}
\end{array}\]
%
To construct a completion of $\Val$, we must choose values for the free variables of $\Val$, ${\Varec{x}}$ and ${\Varec{w_3}}$.
The content of the set $\mathcal{T}^\M_{{\Varec{x}}}( \Val )$ is $\alpha$-equivalent to
\[%\begin{array}{c}
\{
\MU x.\; \const{C}( \const{D}( x )),\;
\const{C}( \MU y.\; \const{D}( y )),\;
\MU x.\; \const{D}( x ),\;
\MU x.\; \const{E}( \const{C}( x )),\;
\MU x.\; \const{C}( \const{E}( x )),\;
\MU x.\; \const{C}( x )
\}
%\end{array}
\]
Consider a model $\M$ that interprets variables in $\Ec$ based on the mapping $\Val$,
for example, $x$ is interpreted as $\Val \ec{x}$, $y_1$ as $\Val \ec{y_1}$ and so on.
Assigning a value for ${\Varec{x}}$ that is $\alpha$-equivalent to a value in $\mathcal{T}^\M_{{\Varec{x}}}( \Val )$
may cause values in the range of $\Val$ to become $\alpha$-equivalent,
which in turn may cause $\M$ to be inconsistent.
For example, say we assign $\MU {\Varec{x}}.\; \const{D}( {\Varec{x}} )$ for ${\Varec{x}}$.
After this substitution, $\Val \ec{y_2}$ is $\MU {\Varec{y_2}}.\; \const{D}( \MU {\Varec{x}}.\; \const{D}( {\Varec{x}} ) )$,
which has normal form $\MU {\Varec{y_2}}.\; \const{D}( {\Varec{y_2}} )$,
which is $\alpha$-equivalent to $\MU {\Varec{x}}.\; \const{D}( {\Varec{x}} )$.
However, this contradicts the disequality $x \not\teq y_2$ in $\Ec$.
Inversely, if the value assigned to ${\Varec{x}}$ is not $\alpha$-equivalent to any term in $\mathcal{T}^\M_{{\Varec{x}}}( \Val )$,
the values in the range of $\Val$ remain $\alpha$-disequivalent.
For example, we may assign values such as 
$\MU {\Varec{x}}.\; \const{E}( {\Varec{x}} )$,
$\MU {\Varec{x}}.\; \const{E}( \MU x_1.\; \const{C}( x_1) )$, or
$\MU {\Varec{x}}.\; \const{D}( \const{C}( {\Varec{x}} ) )$
to ${\Varec{x}}$.
Legal substitutions for $v_{\ec{x}}$ may cause the range of $\Val$ to contain abnormal terms.
After applying the substitution for the third term mentioned above,
$\Val \ec{w_1}$ is $\MU {\Varec{w_1}}.\; \const{C}( \MU {\Varec{x}}.\; \const{D}( \const{C}( {\Varec{x}} ) ) )$,
which has normal form
$\MU {\Varec{w_1}}.\; \const{C}( \MU {\Varec{x}}.\; \const{D}( {\Varec{w_1}} ) )$.
\xend
\end{examplex}

@@

\begin{lemma}
\label{lem:model-completion}
If $\Val$ is constructed for a saturated $\Ec$,
and $\Val^\ast$ is a completion of $\Val$ for normal model $\M$, then
\begin{enumerate}
\item $\Val^\ast \ec{x}$ is $\alpha$-equivalent to a value in $\M( \tau )$, where $\ec{x}$ has type $\tau$,
\item
All subterms $\MU {\Varec{x}}.\; t$ of $\Val^\ast \ec{y}$ are such that
$\Val^\ast \ec{y} \interp{ \MU {\Varec{x}}.\; t } = \Val^\ast \ec{x}$. 
%If $\const{C}( t_1, \ldots, t_n ) \in \ec{x}$,
%then all subterms $\MU v_{\ec{x}}.\; u$ occurring in $\Val^\ast \ec{y}$ 
%are of the form $\MU v_{\ec{x}}.\; \const{C}( w_1, \ldots, w_n )$,
%where 
%and $\Val^\ast \ec{y} \interp{ w_i } = \Val^\ast \ec{t_i}$ for $i = 1, \ldots, n$.
%If $\Val^\ast \ec{x}$ contains a subterm of the form $\MU v_{\ec{x}}.\; \const{C}( w_1, \ldots, w_n )$,
%and $\muvar( w_i ) = v_{\ec{t_i}}$ for $i = 1, \ldots, n$,
%\item
%$t_1 \interp{ t_2 } = \Val^\ast \ec{x}$
%for all $t_1$ with subterm $t_2$ in the range of $\Val^\ast$ where $\muvar(t_2)={\Varec{x}}$,
\item
$\Val^\ast \ec{x} \vsim \Val^\ast \ec{y}$ if and only if $\ec{x} = \ec{y}$.
\end{enumerate}
\end{lemma}
\begin{proof}
To show 1, we first show that $\Val^\ast$ contains no free variables.
Assume by contradiction that $\Val^\ast$ contained a free variable ${\Varec{y}}$ for some $\ec{y}$ of type $\tau$.
Then it must be the case that $\ec{y}$ does not contain a constructor term,
or else ${\Varec{y}}$ would not occur as a free variable in $\Val$.
Consider the case when $\tau$ is finite.
By assumption, $\tau \not\in \Nondata$.
Since \rn{Split} does not apply to $\Ec$, we have $\tau \not\in \Data$.
If $\tau \in \Codata$, then $\tau$ is corecursive by assumption, and by Lemma~\ref{lem:corecursive-singletons},
the cardinality of $\tau$ must be one.
Since \rn{Singleton} does not apply to $\Ec$,
there is only one equivalence class of type $\tau$ in $\Ec$,
and thus there are no terms in $\mathcal{T}^\M_x(\Val^\ast)$ of type $\tau$.
This is a contradiction, since our model completion may assign the value in the domain of $\tau$ to ${\Varec{x}}$.
Now, consider the case when $\tau$ is infinite.
This is also a contradiction,
since there are only a finite number of closed terms in $\mathcal{T}^\M_x(\Val^\ast)$,
and thus our model completion can assign a value not occurring in $\mathcal{T}^\M_x(\Val^\ast)$ to ${\Varec{y}}$.
By construction, $\Val^\ast \ec{x}$ is normal.
Since \rn{Acyclic} does not apply, $\Val \ec{x}$ is acyclic when $\tau \in \Data$.
Moreover, our construction of $\Val^\ast$ applies substitutions of the form
$\{ y \mapsto t \}$, where $t$ is acyclic when the type of $y$ is not a codatatype.
Thus, $\Val^\ast \ec{x}$ is acyclic when $\tau \not\in \Codata$.
Therefore, by definition, $\Val^\ast \ec{x}$ is $\alpha$-equivalent to a value in $\M( \tau )$.

>>>>

\begin{examplex}
Let $\delta$ be a codatatype with the constructors $\const{C}, \const{D}, \const{E} :\delta \rightarrow \delta$.
Let $\Ec$ be the set
$\{
w_1 \teq \const{C}( x ),\;
w_2 \teq \const{E}( w_3 ),\;
x \not\teq y_2,\;
y_1 \teq \const{C}( y_2 ),\;
y_2 \teq \const{D}( x ),\;
z_1 \teq \const{E}( z_2 ),\;
z_2 \teq \const{C}( z_1 )
\}$.
After running the calculus to saturation on $\Ec$, the mapping $\Val$ contains
\[\begin{array}{r@{}c@{}l@{\;}l@{\qquad}r@{}c@{}r@{\;}l@{}l@{\;}l}
\Val \ec{w_1} & {}\,=\,{} & \MU {\Varec{w_1}}. & \const{C}( {\Varec{x}} ) &
\Val \ec{y_1} & {}\,=\,{} & \MU {\Varec{y_1}}. & \const{C}( & \MU {\Varec{y_2}}. & \const{D}( {\Varec{x}} ) ) \\
\Val \ec{w_2} & = & \MU {\Varec{w_2}}. & \const{E}( {\Varec{w_3}} ) &
\Val \ec{y_2} & = & \MU {\Varec{y_2}}. & \multicolumn{2}{@{}l@{}}{\const{D}( {\Varec{x}} )} \\
\Val \ec{w_3} & = & {\Varec{w_3}} &&
\Val \ec{z_1} & = & \MU {\Varec{z_1}}. & \const{E}( & \MU {\Varec{z_2}}. & \const{C}( {\Varec{z_1}} ) ) \\
\Val \ec{x} & = & {\Varec{x}} &&
\Val \ec{z_2} & = & \MU {\Varec{z_2}}. & \const{C}( & \MU {\Varec{z_1}}. & \const{E}( {\Varec{z_2}} ) )
\end{array}\]
%
To construct a completion of $\Val$, we must choose values for the free variables of $\Val$,
namely ${\Varec{w_3}}$ and ${\Varec{x}}$.
The content of the set \smash{$\mathcal{T}^\M_{{\Varec{x}}}( \Val )$} is $\alpha$-equivalent to
\[%\begin{array}{c}
\{
\MU x.\; \const{C}( x ),\;
\MU x.\; \const{C}( \const{D}( x )),\;
\const{C}( \MU x.\; \const{D}( x )),\;
\MU x.\; \const{C}( \const{E}( x )),\;
\MU x.\; \const{D}( x ),\;
\MU x.\; \const{E}( \const{C}( x ))
\}
%\end{array}
\]
Consider a model $\M$ that interprets variables in $\Ec$ based on the mapping $\Val$:
$x$ is interpreted as $\Val \ec{x}$, $y_1$ as $\Val \ec{y_1}$, and so on.
Assigning a value for ${\Varec{x}}$ that is $\alpha$-equivalent to a value in $\mathcal{T}^\M_{{\Varec{x}}}( \Val )$
may cause values in the range of $\Val$ to become $\alpha$-equivalent,
which in turn may cause $\M$ to be inconsistent.
For example, assign $\MU {\Varec{x}}.\; \const{D}( {\Varec{x}} )$ for ${\Varec{x}}$.
After this substitution, $\Val \ec{y_2}$ is $\MU {\Varec{y_2}}.\; \const{D}( \MU {\Varec{x}}.\; \const{D}( {\Varec{x}} ) )$,
which has normal form $\MU {\Varec{y_2}}.\; \const{D}( {\Varec{y_2}} )$,
which is $\alpha$-equivalent to $\MU {\Varec{x}}.\; \const{D}( {\Varec{x}} )$.
However, this contradicts the disequality $x \not\teq y_2$ in~$\Ec$.
On the other hand, if the value assigned to ${\Varec{x}}$ is not $\alpha$-equivalent to any term in $\mathcal{T}^\M_{{\Varec{x}}}( \Val )$,
the values in the range of $\Val$ remain $\alpha$-disequivalent.
For example, we may assign values such as 
$\MU {\Varec{x}}.\; \const{E}( {\Varec{x}} )$,
$\MU {\Varec{x}}.\; \const{E}( \MU x_1.\; \const{C}( x_1) )$, or
$\MU {\Varec{x}}.\; \const{D}( \const{C}( {\Varec{x}} ) )$
to ${\Varec{x}}$.
Legal substitutions for $v_{\ec{x}}$ may cause the range of $\Val$ to contain abnormal terms.
After applying the substitution for the third term mentioned above,
$\Val \ec{w_1}$ is $\MU {\Varec{w_1}}.\; \const{C}( \MU {\Varec{x}}.\; \const{D}( \const{C}( {\Varec{x}} ) ) )$,
which has normal form
$\MU {\Varec{w_1}}.\; \const{C}( \MU {\Varec{x}}.\; \const{D}( {\Varec{w_1}} ) )$.
\xend
\end{examplex}

@@

\begin{lemma}
\label{lem:model-completion}
If $\Val$ is constructed for a saturated set $\Ec$
and $\Val^\ast$ is a completion of $\Val$ for normal model $\M$, then
\begin{enumerate}
\item $\Val^\ast \ec{x}$ is $\alpha$-equivalent to a value in $\M( \tau )$, where $\ec{x}$ has type $\tau$.
\item
All subterms $\MU {\Varec{x}}.\; t$ of $\Val^\ast \ec{y}$ are such that
$\Val^\ast \ec{y} \interp{ \MU {\Varec{x}}.\; t } = \Val^\ast \ec{x}$. 
%If $\const{C}( t_1, \ldots, t_n ) \in \ec{x}$,
%then all subterms $\MU v_{\ec{x}}.\; u$ occurring in $\Val^\ast \ec{y}$ 
%are of the form $\MU v_{\ec{x}}.\; \const{C}( w_1, \ldots, w_n )$,
%where 
%and $\Val^\ast \ec{y} \interp{ w_i } = \Val^\ast \ec{t_i}$ for $i = 1, \ldots, n$.
%If $\Val^\ast \ec{x}$ contains a subterm of the form $\MU v_{\ec{x}}.\; \const{C}( w_1, \ldots, w_n )$,
%and $\muvar( w_i ) = v_{\ec{t_i}}$ for $i = 1, \ldots, n$,
%\item
%$t_1 \interp{ t_2 } = \Val^\ast \ec{x}$
%for all $t_1$ with subterm $t_2$ in the range of $\Val^\ast$ where $\muvar(t_2)={\Varec{x}}$,
\item
$\Val^\ast \ec{x} \vsim \Val^\ast \ec{y}$ if and only if $\ec{x} = \ec{y}$.
\end{enumerate}
\end{lemma}
\begin{proof}
To show 1, we first show that $\Val^\ast$ contains no free variables.
Assume by contradiction that $\Val^\ast$ contained a free variable ${\Varec{y}}$ for some $\ec{y}$ of type $\tau$.
Then it must be the case that $\ec{y}$ does not contain a constructor term,
or else ${\Varec{y}}$ would not occur as a free variable in $\Val$.
Consider the case when $\tau$ is finite.
By assumption, $\tau \not\in \Nondata$.
Since \rn{Split} does not apply to $\Ec$, we have $\tau \not\in \Data$.
If $\tau \in \Codata$, then $\tau$ is corecursive by assumption, and by Lemma~\ref{lem:corecursive-singletons},
the cardinality of $\tau$ must be one.
Since \rn{Singleton} does not apply to $\Ec$,
there is only one equivalence class of type $\tau$ in $\Ec$,
and thus there are no terms in $\mathcal{T}^\M_x(\Val^\ast)$ of type $\tau$.
This is a contradiction, since our model completion may assign the value in the domain of $\tau$ to ${\Varec{x}}$.
Now, consider the case when $\tau$ is infinite.
This is also a contradiction,
since there are only a finite number of closed terms in $\mathcal{T}^\M_x(\Val^\ast)$,
and thus our model completion can assign a value not occurring in $\mathcal{T}^\M_x(\Val^\ast)$ to ${\Varec{y}}$.
By construction, $\Val^\ast \ec{x}$ is normal.
Since \rn{Acyclic} does not apply, $\Val \ec{x}$ is acyclic when $\tau \in \Data$.
Moreover, our construction of $\Val^\ast$ applies substitutions of the form
$\{ y \mapsto t \}$, where $t$ is acyclic when the type of $y$ is not a codatatype.
Thus, $\Val^\ast \ec{x}$ is acyclic when $\tau \not\in \Codata$.
Therefore, by definition, $\Val^\ast \ec{x}$ is $\alpha$-equivalent to a value in $\M( \tau )$.
